---
title: "ENVS 193DS Final Project: Final Effective Data Visualization and Communication"
author: "Siaa Singh"
date: "06/10/2025"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,   # suppress library startup messages
  warning = FALSE    # suppress warnings
)
```




```{r}
# Load required libraries
library(tidyverse)
library(readr)
library(ggplot2)
library(dplyr)

# Read in the SST dataset
sst <- read_csv("../data/SST_update2023.csv")
```

Problem 1 
a. a. Transparent Statistical Methods
Part 1: The author used a correlation test, likely Pearson’s correlation, to evaluate the 
relationship between distance from the headwater and total nitrogen load. This is implied by the phrasing “correlation” and the numeric p-value. The author used an Analysis of Variance (ANOVA) to 
compare the mean nitrogen loads between different sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands), as indicated by the reference to 
“average nitrogen load” and the overall p-value for group differences.

b. Including means and standard deviations for nitrogen load by each source category would give readers a clearer picture of how nitrogen contributions differ across land use types.Providing the number of observations per nitrogen source would help readers understand the reliability of the ANOVA results. For example, if one group had a very small sample size, its contribution to the test outcome might be questionable.

c. Nitrogen load increased with distance from the headwater, suggesting a spatial accumulation effect along the river. A correlation test was used to assess this relationship 
(test: Pearson’s correlation; r = [correlation coefficient], p = 0.03, α = 0.05).
There were statistically significant differences in nitrogen load among the five source categories, suggesting that some sources contribute disproportionately to nitrogen levels in the delta. 
An ANOVA was used to compare group means (test: ANOVA; F = [test statistic], df = 
[degrees of freedom], p = 0.02, α = 0.05).


2 a. 

```{r}
# Load required libraries
library(tidyverse)
library(lubridate)

#  Read in the SST data
sst <- read_csv("../data/SST_update2023.csv")

# Clean and summarize the SST data
sst_clean <- sst %>%
  # Extract year and month from the date column
  mutate(
    year = year(date),
    month = month(date, label = TRUE, abbr = TRUE)
  ) %>%
  # Group by year and month to calculate monthly means
  group_by(year, month) %>%
  summarise(
    mean_monthly_sst = mean(temp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Filter for years 2018 to 2023
  filter(year %in% 2018:2023)

# View a random sample of 5 rows
slice_sample(sst_clean, n = 5)

# Display the structure of the cleaned data
str(sst_clean)

```

2. b 
```{r}
# Filter data for the years of interest
sst_filtered <- sst_clean %>%
  filter(year %in% 2018:2023)

# Create the plot
ggplot(sst_filtered, aes(x = month, y = mean_monthly_sst, group
                         = year, color = year)) +
  # Add lines for each year
  geom_line(size = 1) +
  # Add points for each month
  geom_point(size = 2) +
  # Use a gradient scale: light to dark
  scale_color_gradient(low = "lightblue", high = "navy") +
  # Label the axes
  labs(
    x = "Month",
    y = "Mean monthly sea surface temperature (°C)",
    color = "Year"
  ) +
  # Put legend inside the plot
  theme(
    legend.position = c(0.15, 0.75),  
    # top left-ish
    legend.background = element_rect(fill = "white", color = "black"),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA)
  )

```

Problem 3. 

```{r}
# Load required libraries
library(tidyverse)  # For data wrangling and ggplot
library(DHARMa)     # For checking model diagnostics
library(MuMIn)      # For AIC model comparison
library(broom)      # For tidying model output (optional but useful)

```

```{r}
nest <- read_csv("../data/occdist.csv")
```

a. In this dataset, the "sp" variable represents Swift Parrot occupancy.
1 indicates that the parrot used a nest box during the season. The value of 
0 indicates that a box was not used/occupied. 

b. The main distinction is that the nest box intervention's main conservation goal is 
for Swift Parrots, a species that is critically endangered. 
Swift Parrot availability may be lowered by non-target or competing species 
like Common Starlings and Tree Martins, which may also inhabit nest boxes.

c. Nest box occupancy was measured in 2016 and 2019, which are referred to as the two "seasons" in this study. Researchers can determine whether temporal variation influences nest box use by comparing the environmental conditions and Swift Parrot occupancy levels in these years.

d. 
```{r}
# Model table for Problem 3d
library(tibble)

model_table <- tibble::tibble(
  `Model #` = c(1, 2, 3, 4),
  `Season Included` = c("No", "Yes", "Yes", "No"),
  `Distance to Edge Included` = c("No", "Yes", "No", "Yes"),
  `Model Description` = c(
    "Null model — assumes constant occupancy",
    "Saturated model — includes both season and distance",
    "Season-only model",
    "Distance-only model"
  )
)

# View the model table
model_table

```

e. 
```{r}
# Run all logistic regression models for Swift Parrot occupancy

# Null model – no predictors
model1 <- glm(sp ~ 1, data = nest, family = binomial)

# Saturated model – includes both season and distance
model2 <- glm(sp ~ season + `edge distance`, data = nest, family = binomial)

# Season-only model
model3 <- glm(sp ~ season, data = nest, family = binomial)

# Distance-only model
model4 <- glm(sp ~ `edge distance`, data = nest, family = binomial)

```

3f 
```{r}
# Load DHARMa (only if not already loaded)
library(DHARMa)

# Simulate residuals for each model
sim1 <- simulateResiduals(model1)
sim2 <- simulateResiduals(model2)
sim3 <- simulateResiduals(model3)
sim4 <- simulateResiduals(model4)

# Plot residuals for each model
plot(sim1, main = "Model 1: Null Model")
plot(sim2, main = "Model 2: Saturated Model")
plot(sim3, main = "Model 3: Season-only Model")
plot(sim4, main = "Model 4: Distance-only Model")

```

3g. 
```{r}
# Load MuMIn if not already
library(MuMIn)

# Compare models using AIC
model_selection <- AIC(model1, model2, model3, model4)

# Print AIC table (this is part of your output)
model_selection

```

The best model as determined by Akaike’s Information Criterion (AIC) was the saturated model 
(model 2), which included both season and distance from forest edge as predictors of Swift 
Parrot nest box occupancy. This model had the lowest AIC value (226.21), indicating it 
provides the best trade-off between goodness of fit and model complexity among those tested.

3f. 
```{r}
# Load required packages
library(ggplot2)
library(dplyr)

# Create a new dataset for predictions across edge distances in both seasons
prediction_data <- expand.grid(
  `edge distance` = seq(min(nest$`edge distance`), max(nest$`edge distance`), length.out = 100),
  season = unique(nest$season)
)

# Generate predictions with confidence intervals using model2
predictions <- predict(model2, newdata = prediction_data, type = "response", se.fit = TRUE)

# Bind predictions to the prediction_data
prediction_data$fit <- predictions$fit
prediction_data$lower <- predictions$fit - 1.96 * predictions$se.fit
prediction_data$upper <- predictions$fit + 1.96 * predictions$se.fit

# Plot model predictions + underlying data
ggplot() +
  # Add raw occupancy data
  geom_jitter(data = nest, aes(x = `edge distance`, y = sp, color = as.factor(season)), width = 15, height = 0.03, alpha = 0.6) +
  
  # Add prediction line
  geom_line(data = prediction_data, aes(x = `edge distance`, y = fit, color = as.factor(season)), size = 1.2) +
  
  # Add 95% confidence ribbon
  geom_ribbon(data = prediction_data, aes(x = `edge distance`, ymin = lower, ymax = upper, fill = as.factor(season)), alpha = 0.2) +
  
  # Customize labels and appearance
  labs(
    x = "Distance from Forest Edge (meters)",
    y = "Probability of Swift Parrot Nest Box Occupancy",
    color = "Season",
    fill = "Season",
    title = "Predicted Probability of Swift Parrot Nest Box Occupancy by Distance from Forest Edge"
  ) +
  theme_minimal() +                                # Clean theme
  theme(
    panel.grid = element_blank(),                  # Remove gridlines
    plot.title = element_text(hjust = 0.5)         # Center title
  ) +
  scale_color_manual(values = c("darkorange", "steelblue")) +
  scale_fill_manual(values = c("darkorange", "steelblue"))

```

3i. 
This figure displays the model-predicted probability of Swift Parrot (Lathamus discolor) 
nest box occupancy (solid lines) as a function of distance from the forest edge (in meters), 
for the two study seasons: 2016 and 2019. Observed occupancy values (1 = occupied, 0 = unoccupied) 
are jittered in the background as points. Shaded ribbons represent 95% confidence intervals 
around predictions. The model suggests that probability of occupancy declines with increasing 
distance from the forest edge and differs slightly by season.
Data source: Stojanovic et al. (2021), Dryad, https://doi.org/10.5061/dryad.83bk3j9sb

3j. 
```{r}
# Create new data frame for prediction
prediction_data <- expand.grid(
  `edge distance` = c(0, 900),
  season = c(2016, 2019)
)

# Predict using the best model (model2)
# Set type = "response" to get probabilities
predicted_probs <- predict(model2, newdata = prediction_data, type = "response", se.fit = TRUE)

# Combine predictions with original new data
prediction_results <- prediction_data %>%
  mutate(
    predicted_prob = predicted_probs$fit,
    lower_CI = predicted_probs$fit - 1.96 * predicted_probs$se.fit,
    upper_CI = predicted_probs$fit + 1.96 * predicted_probs$se.fit
  )

# Show the final prediction table
prediction_results

```

3k. 
Based on our results, Swift Parrot nest box occupancy was highest at the forest edge (0 m) in both seasons. A predicted probability of 0.48 in 2016 and 0.30 in 2019. 
At 900 m from the edge, predicted occupancy declined sharply to 0.12 in 2016 and 0.06 in 2019. 
The figure from part h supports this trend, showing a clear decrease in occupancy with increasing distance, and slightly higher occupancy rates across all distances in 2016. This pattern suggests 
that proximity to forest edge provides ecological advantages—such as better food availability or nesting habitat—which may be critical for Swift Parrots. The stronger edge effect in 2016 may also reflect variation in habitat quality or population size between the two years. 


 
4a. Comparing Visualizations

1. How are the visualizations different from each other in the way you have represented your data?
The affective visualization (HW 3) uses color to represent mood and emphasizes emotional impact, with bold colors and labels to highlight personal experience. In contrast, the exploratory visualization
(HW 2) presents the data more neutrally and statistically, using a boxplot to summarize the distribution of hours outdoors across homework load categories.

2. What similarities do you see between all your visualizations?
Both visualizations explore the relationship between daily behavior (time outdoors) and another variable—either emotional state or workload. They also use clear axes and labeling to 
communicate their message effectively, and both make categorical comparisons (day/mood vs. 
homework load).

3. What patterns do you see in each visualization? Are these different between visualizations? 
If so, why?
In the effective plot, I see that days labeled "Happy" are associated with higher average outdoor hours, suggesting a potential link between mood and outdoor activity. In the exploratory plot, 
"Low" homework days have a higher median outdoor time compared to "High" or "Medium" workload 
days. Both show a positive relationship between free time and time spent outdoors, 
reinforcing each other even though they’re framed differently.

4. What kinds of feedback did you get during week 9 in workshop or from the instructors?
I was encouraged to simplify the mood categories and adjust the annotation placement for clarity. 
I implemented this by reducing the number of moods to three and repositioning the labels so they wouldn’t overlap with the bars. I also added a bolded title and emphasized the emotional 
framing, per suggestions to make the story more engaging.

